Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2023-11-17 SEIR.tex
Edit figures, talk about structured populations

This chapter uses mechanistic models for estimating the transmission of SARS-CoV-2.
This is an alternative approach to the statistical methods used in \cref{E-backcalc}.
Mechanistic models explicitly represent the infection process driving the spread of infectious disease.
The explicit representation means that a mechanistic model's state and parameters have biological and/or epidemiological interpretations; these interpretations provide scientific understanding of the disease.
Of particular interest is that the basic reproduction number, $R_0$, and the time-varying reproduction number, $R_t$, can both be calculated as simple functions of the model parameters; in \cref{E-backcalc} additional modelling is required to calculate these numbers.
Furthermore, mechanistic models can be used for scenario-based modelling, whereby the effect of different interventions can be simulated.
All of these properties make mechanistic models a useful tool for understanding and controlling infectious diseases.
However, they make stronger assumptions about the disease than statistical models, and often perform worse in short-term forecasting\todo{add a citation}.

@@

Notation in this chapter differs from the rest of this thesis.
I adopt conventions and notation from the infectious disease modelling field, although this is not consistent in the literature.
In particular, upper-case letters denote numbers of people, and lower-case letters denote proportions.
Whether a variable is random or not should be clear from the context, or will be explicitly stated if needed.

@@

This section introduces the class of mechanistic transmission models known as \emph{compartmental} models.
Compartmental models are the most widely-used class of mechanistic infectious disease model.
A compartmental model is named because they place every member of the population under study into a compartment; each compartment represents a disease state (\eg susceptible).

A compartmental model consists of a state vector $\vec{x}(t)$ and a set of parameters $\vec{\theta}$.
The elements of $\vec{x}(t)$ are the proportion of the population (or number of individuals) in each compartment at time $t$.
The parameters $\vec{\theta}$ are constants that determine, along with the model structure, how the state vector changes over time.
Compartmental models can be \emph{deterministic} or \emph{stochastic}.
Deterministic models are defined by the state vector $\vec{x}(t)$ being a deterministic, although rarely analytically-tractable, function of the parameters $\vec{\theta}$ and the initial state $\vec{x}(0)$.
%The transmission model corresponds to the infection process (see \cref{E-inc-prev:sec:infection-process}).
%There remains randomness in the observation model (described in \cref{SEIR:sec:observation}).
Stochastic models specify a probability distribution $p(\vec{x} \mid \vec{\theta})$ for the state vector $\vec{x}(t)$ at time $t$.
In general, this means, there is a dependence between $\vec{x}(t)$ and $\vec{x}(t')$.
This relaxes the assumption of independent increments in the counting process formulation of the infection process (see \cref{E-inc-prev:sec:infection-process}).

Deterministic models are also known as \emph{mean-field} models because they can are the behaviour corresponding to a stochastic model's mean.
When the number of infected individuals is large, then this mean-field behaviour is a good approximation of the fully stochastic model, due to the law of large numbers\todo{ref mean-field and approximation error}.
An alternative justification, when making noisy observations of the system, is that the additional noise from the stochastic model is negligible compared to this observation noise.
This argument is similar to the justification for neglecting the stochasticity of the prevalence process in \cref{E-inc-prev:sec:observation-process}.
The remainder of this chapter uses only deterministic models, because throughout the time period this thesis considers there were enough infectious individuals.
For an introduction to stochastic models, see \textcite[chapter 6]{keelingModeling}.
% \todo{Maybe say something about still having a stochastic observation model, if not clear enough from intro}

@@

\Cref{SEIR:sec:SIR} begins by introducing the \emph{susceptible-infected-recovered} (SIR) model, a simple compartmental model for a disease which confers immunity.
The SIR model is named because each individual in the population is divided into one of three states: susceptible, infectious, or recovered.
This model has several assumptions which make it inappropriate to model SARS-CoV-2; therefore, I introduce extensions to relax these assumptions.
First, \cref{SEIR:sec:non-exponential} relaxes the assumption of an exponentially-distributed infectious period to a gamma distribution.
Then, \cref{SEIR:sec:SEIR} relaxes the assumption that individuals are infectious immediately after they are infected; it does this by extending the SIR model to the SEIR model, including an \emph{exposed} compartment.
Next, \cref{SEIR:sec:structured-populations} relaxes the homogeneous population assumption, introducing stratification by important characteristic(s).
Finally, \cref{SEIR:sec:time-varying-foi} relaxes the assumption of a constant transmission rate to allow for time-varying behavioural changes.

@@

\begin{figure}[h]
\centering
\begin{tikzpicture}[
    node distance = 2.5cm,
    on grid,
    auto,
    ->,>=stealth',
    every state/.style={draw,rectangle,very thick},
    ]

@@

    \path (S) edge node {$\beta i$} (I)
          (I) edge node {$\gamma$} (R);
\end{tikzpicture}
  \caption[The SIR model]{Schematic of the basic SIR model.}
  \label{SEIR:fig:SIR}
\end{figure}

The simplest compartmental model for a disease which confers immunity is the (SIR) model\todo{ref for SIR being simplest}.
The name refers to the three compartments in the model, displayed in \cref{SEIR:fig:SIR}.
Individuals in the susceptible compartment can be infected; $s(t)$ is the proportion of the population that is susceptible at time $t$.
Individuals in the infectious compartment can infect others, and eventually recover; $i(t)$ is the proportion of the population that is infectious at time $t$.
Individuals in the recovered compartment are immune to the disease, and cannot be infected again; $t(t)$ is the proportion of the population that have recovered by time $t$.
The model's state is $\vec{x}(t) = (s(t), i(t), r(t))^T$.

The SIR model is described by the following system of ODEs~\autocite[19]{keelingModeling}:
\begin{align}
\frac{ds(t)}{dt} &= -\lambda(t) s(t) \\
\frac{di(t)}{dt} &= \lambda(t) s(t) - \gamma i(t) \\
\frac{dr(t)}{dt} &= \gamma i(t)
\end{align}
where $\lambda(t) = \beta i(t)$, is the \emph{force of infection}: the risk of infection for a susceptible individual at time $t$~\autocite[17]{keelingModeling}.
Two parameters are present in this model: $\beta$ and $\gamma$, \ie $\vec{\theta} = (\beta, \gamma)^T$.
$\beta$ is the transmission rate, absorbing several terms as described below.
%\todo{check what beta is called: effective transmission rate or transmission rate or something else?}, or the number of individuals an infected individual infects per day.
$\gamma$ is the recovery rate, or the proportion of infected individuals that recover per day; hence, $1/\gamma$ is the mean infectious period~\autocite[367]{keelingModeling}.
These parameters are assumed to be constant (an assumption relaxed in \cref{SEIR:sec:time-varying-foi}).

The term $\beta s(t)i(t)$ is the rate at which susceptible individuals (as a proportion of the total population size) are infected~\autocite[18]{keelingModeling}.
Susceptible individuals move from the $s(t)$ to $i(t)$ when they are infected.
This term can be arrived at by considering a single infected individual~\autocite[214]{kretzschmarMathematical}.
Assume this individual comes into contact with other individuals in the population at a rate of $\kappa$ per day.
Further, assume the population is \emph{well-mixed}, meaning that each contact is with an individual chosen uniformly at random from the population.
For such a randomly selected individual, the probability that they are susceptible is $s(t)$.
Each of the infected individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
Therefore, in expectation, the infected individual makes contact with susceptible individuals at a rate of $\kappa s(t)$.
Assume that there is a constant probability of transmission upon contact between an infected and susceptible individual, $q$.
Combined with the above, the expected rate of infections caused by this infected individual is $\kappa s q$.
The two constant terms here (number of contacts and probability of transmission) are absorbed into the transmission rate, $\beta = \kappa q$.
Therefore, the rate of new infections generated by a single infectious individual is $\beta s(t)$.
Multiplying by the number of infected individuals and dividing by the population size gives the rate at which susceptible individuals are infected as a proportion of the population, $\beta s(t) i(t)$.

The term $\gamma i(t)$, the rate at which infected individuals (as a proportion of the total population size) recover, follows directly from the definitions of $\gamma$ and $i(t)$.

A few properties and assumptions of the SIR model are worth noting.
\begin{itemize}
    \item The basic reproduction number (discussed previously in\todo{ref intro}) for this model is $R_0 = \beta / \gamma$~\autocite[20]{keelingModeling}.
    Intuitively, this can be thought of as the average number of infections caused by an infectious individual per day, $\beta$, multiplied by the average number of days they are infectious for, $1/\gamma$.
    \item The population is closed.
    That is, the total population size is constant, and no infections are imported from outside the population.
    Births and deaths of individuals are also excluded, although they can easily be incorporated in situations where they are relevant~\autocites[26]{keelingModeling}[214]{kretzschmarMathematical}.
    \item There is no analytical solution to the system of ODEs~\autocite[25]{keelingModeling}: numerical methods are required to solve the system.
    \item In addition to the parameters, the state of the system at some time (normally the initial state, the state at time $t=0$) is required to solve the system.
    \item Immunity is assumed to last until the end of the modelled period.
    For many diseases, including SARS-CoV-2, this is a reasonable assumption as long as the modelled period is short, as it will be in this chapter\todo{ref that SARS-CoV-2 immunity is long enough to not be an issue}.
    For modelling longer time periods, waning immunity can be incorporated\todo{ref waning immunity models}.
\end{itemize}

\subsection{Non-exponential waiting times} \label{SEIR:sec:non-exponential}
\begin{figure}[h]
\begin{tikzpicture}[
    node distance = 2.5cm,
    on grid,
    auto,
    ->,>=stealth',
    every state/.style={draw,rectangle,very thick},
    ]

@@

    \path (S) edge node {$\beta i$} (I1)
          (I1) edge node {$n\gamma$} (I2)
          (I2) edge node {$n\gamma$} (I3)
          (I3) edge node {$n\gamma$} (I4)
          (I4) edge node {$n\gamma$} (R);
\end{tikzpicture}
  \caption[The SIR model with non-exponential infectious period]{Schematic of the SIR model with non-exponential infectious period. The infectious period is modified by having multiple I states.}
  \label{SEIR:fig:SIR-gamma}
\end{figure}

The basic SIR model implies that the distribution of the time spent in the infected compartment is exponential~\autocite[96]{keelingModeling}.
This is an unrealistic assumption for most diseases and leads to an underestimation of the reproduction number if the model is fit to data~\autocites{lloydRealistic}{wearingAppropriate}.
A standard extension to the SIR model is to allow the infectious period to follow a gamma distribution~\autocite[94]{keelingModeling}.
This is done by adding multiple infected compartments (see \cref{SEIR:fig:SIR-gamma}).
The system can now be written as follows:
\begin{align}
\frac{ds(t)}{dt} &= -\lambda(t) s(t)\\
\frac{di_1(t)}{dt} &= \lambda(t) s(t) - n\gamma i_1(t) \\
\frac{di_2(t)}{dt} &= n\gamma i_1(t) - n \gamma i_2(t) \\
&\vdots \nonumber \\
\frac{di_n(t)}{dt} &= n\gamma i_{n-1}(t) - n \gamma i_n(t) \\
\frac{dr(t)}{dt} &= n\gamma i_n(t)
\end{align}
where $n$ is the number of infected compartments, and redefining $\lambda(t) = \beta \sum_{j=1}^n i_j(t)$ 
In this model the infectious period is distributed $\GamDist(n, n\gamma)$, a realistic model for a variety of diseases~\autocite{wearingAppropriate}.
This parameterisation is chosen to give $1/\gamma$ as the mean infectious period (the gamma distribution and its expectation is defined in \cref{E-distributions}).
% (it is slightly unclear what it means for a variable to have a distribution in a deterministic model, the clearest explanation is that this is the distribution which the ).
A gamma distribution with the first parameter equal to 1 is an exponential distribution, connecting this model to the basic SIR model in \cref{SEIR:sec:SIR}.
The distribution for the time spent in the infected compartments tends to a constant (\ie the same for all individuals) as $n \to \infty$~\autocite{lloydRealistic}.
The reproduction number in terms of the parameters is unchanged by this extension, \ie $R_0 = \beta / \gamma$.

@@

\begin{figure}[h]
\begin{tikzpicture}[
    node distance = 2cm,
    on grid,
    auto,
    ->,>=stealth',
    every state/.style={draw,rectangle,very thick},
    remember picture, overlay
    ]

    \node[state, xshift=-2cm] (S) {$s$};
    \node[state, right=of S] (E1) {$e_1$};
    \node[state, right=of E1] (E2) {$e_2$};
    \node[state, right=of E2, draw=none] (E3) {$\cdots$};
    \node[state, right=of E3] (E4) {$e_m$};
    \node[state, right=of E4] (I1) {$i_1$};
    \node[state, right=of I1] (I2) {$i_2$};
    \node[state, right=of I2, draw=none] (I3) {$\cdots$};
    \node[state, right=of I3] (I4) {$i_n$};
    \node[state, right=of I4] (R) {$r$};

    \path (S) edge node {$\beta i$} (E1)
          (E1) edge node {$m\sigma$} (E2)
          (E2) edge node {$m\sigma$} (E3)
          (E3) edge node {$m\sigma$} (E4)
          (E4) edge node {$m\sigma$} (I1)
          (I1) edge node {$n\gamma$} (I2)
          (I2) edge node {$n\gamma$} (I3)
          (I3) edge node {$n\gamma$} (I4)
          (I4) edge node {$n\gamma$} (R);
\end{tikzpicture}
  \caption[The SEIR model]{Schematic of the SEIR model.}
  \label{SEIR:fig:SEIR}
\end{figure}

The SIR model assumes that individuals are immediately infectious upon infection.
This is unrealistic for most diseases, as there is a period between an individual being infected and them being infectious, known as the \emph{latent period}.
\todo{perhaps link with where define phases of infection in background chapters}
Ignoring the latent period will again overestimate the reproduction number~\autocite{wearingAppropriate}.

@@

To relax the assumption of a homogeneous population, the population can be stratified.
If the strata have differing transmission patterns, this can change the dynamics of the epidemic.
The different patterns could be due to behavioural factors (\eg high-numbers of sexual partners in a sexually-transmitted disease~\autocite[69]{keelingModeling} or younger age groups making many more contacts for respiratory diseases~\autocite[176]{andersonInfectious}) or biological factors (\eg genetics reducing the susceptibility of some individuals~\autocite[208]{andersonInfectious}).
I will assume that the disease progression (\eg length of the exposed and infectious periods) is the same for all strata, although relaxing this assumption is easy.
Stratification requires two major changes model changes: extension of the state vector and describing the between-strata interactions.

To incorporate the stratification into the model, first I introduce some additional notation.
Denote the set of strata of size $A$, as $\set{A}$.
A stratum will be denoted by $a \in \set{A}$ whose total population is $N_a$.

First, the state vector is now a vector of vectors, one for each of the compartments.
For example, in the SEIR model of \cref{SEIR:sec:SEIR}, $\vec{x} = (\vec{s}, \vec{e_1}, \dots, \vec{e_m}, \vec{i_1}, \dots, \vec{i_n}, \vec{r})^T$.
The $a$th element of each vector is the proportion of strata $a$ in that compartment.
For example, $\vec{s} = (s_1, \dots s_A)^T$ where $s_a$ is the proportion of the $a$th strata that are susceptible.
Hence, $N_a s_a$ is the number of susceptible people in stratum $a$.
These vectors simply replace the scalars in the previous equations.
Various other definitions are possible and equivalent, for example defining $s_a$ as the proportion of the total population that are both in strata $a$ and susceptible or parameterised using numbers of individuals\todo{cite Keeling}.

Second, the interactions between strata must be described.
This is done by replacing the scalar $\beta$ with a \emph{Who Acquires Infection From Whom} (WAIFW) matrix, $\matr{\beta}$.
The force of infection is now a vector, $\vec{\lambda} = \matr{\beta} \sum_{j=1}^n \vec{i_j}$.
The matrix element $\beta_{aa'}$ is the rate at which individuals in stratum $a$ infect individuals in stratum $a'$.
In practical situations, estimating all of these parameters is infeasible, so some assumptions must be made to reduce the number of parameters.
Common assumption include assuming symmetry, \ie $\beta_{aa'} = \beta{a'a}$ or a decomposable independence, \ie $\beta_{aa'} = \beta \beta^\text{inf}_a \beta^\text{susc}_{a'}$~\todo{ref A+M on these parameterisations}.
Alternatively, as in the scalar case, $\beta_{aa'}$ can be decomposed into the contact rate between strata $a$ and $a'$, $\kappa_{aa'}$ and the probability of transmission upon contact, $q_{aa'}$, and then assumptions made about the components.

\todo[inline]{start with a generic intro on structured populations and use of matrices, then give age as a motivating example, finally specifics go elsehwere}

\todo[inline]{Possibly add a paragraph about the importance of age and mixing patterns, maybe show a contact matrix?}

To allow for age, the population is divided into age groups.
The state of each compartment is now a vector.
The proportion of the population that is susceptible is $\vec{s} = (s_1, \dots, s_A)$ where $s_a$ is the proportion of the $a$th age group that is susceptible.
Similarly, for the other compartments.

The modelling approach here can be used to stratify the population by any characteristic.
For example, HIV/AIDS models incorporated subgroups with different patterns of sexual activity~\autocite{kretzschmarMathematical}.
In other contexts, transmission is strongly influenced by socio-economic status, and the model can be stratified by income levels~\autocite{menkirIncorporating}.
In this case, $i$ simply represents the $i$th strata.

The assumption of a well-mixed population is relaxed by using a contact matrix.
Now, only conditional on age are all individuals identical.

>>>>

This chapter uses mechanistic models for estimating the transmission of SARS-CoV-2.
This is an alternative approach to the statistical methods used in \cref{E-backcalc}.
Mechanistic models explicitly represent the infection process driving the spread of infectious disease.
The explicit representation means that a mechanistic model's state and parameters have biological and/or epidemiological interpretations; these interpretations provide scientific understanding of the disease.
Of particular interest is that the basic reproduction number, $R_0$, and the effective reproduction number, $R_e(t)$, can both be calculated as simple functions of the model parameters; using \cref{E-backcalc} additional modelling would be required to estimate these numbers.
Furthermore, mechanistic models can be used for scenario-based modelling, whereby the effect of different interventions can be simulated.
All of these properties make mechanistic models a useful tool for understanding and controlling infectious diseases.
However, they make stronger assumptions about the disease than statistical models, and often perform worse in short-term forecasting\todo{add a citation}.

@@

Notation in this chapter differs from the rest of this thesis.
I adopt conventions and notation from the infectious disease modelling field, although this is not consistent in the literature.
In particular, case no longer signifies whether a variable is random or not, but will be clear from the context.

@@

\todo[inline]{ref intro to compartmental model setup}
This section introduces the class of mechanistic transmission models known as \emph{compartmental} models.
Compartmental models are the most widely-used class of mechanistic infectious disease model.
A compartmental model is named because they place every member of the population under study into a compartment; each compartment represents a disease state (\eg susceptible).

A compartmental model consists of a state vector $\vec{x}(t)$ and a set of parameters $\vec{\theta}$.
The elements of $\vec{x}(t)$ are the proportion of the population (or number of individuals) in each compartment at time $t$.
The parameters $\vec{\theta}$ are constants that determine, along with the model structure, how the state vector changes over time.
Compartmental models can be \emph{deterministic} or \emph{stochastic}.
Deterministic models are defined by the state vector $\vec{x}(t)$ being a deterministic, although rarely analytically-tractable, function of the parameters $\vec{\theta}$ and the initial state $\vec{x}(0)$.
%The transmission model corresponds to the infection process (see \cref{E-inc-prev:sec:infection-process}).
%There remains randomness in the observation model (described in \cref{SEIR:sec:observation}).
Stochastic models are most commonly specified in discrete time, using a Markov assumption.
That is, they specify a probability distribution $p(\vec{x}(t) \mid \vec{x}(t-1), \vec{\theta})$.
This relaxes the assumption of independent increments in the counting process formulation of the infection process (see \cref{E-inc-prev:sec:infection-process}).

Deterministic models are also known as \emph{mean-field} models because they can are the behaviour corresponding to a stochastic model's mean.
When the number of infected individuals is large, the law of large numbers means that the mean-field behaviour is a good approximation of the fully stochastic model~\autocite[20]{diekmannMathematical}.
An alternative justification, when making noisy observations of the system, is that the additional noise from the stochastic model is negligible compared to this observation noise.
This second justification is of the same form as the one made for neglecting the stochasticity of the prevalence process in \cref{E-inc-prev:sec:observation-process}.
The remainder of this chapter uses only deterministic models, because throughout the time period this thesis considers there were enough infectious individuals.
For an introduction to stochastic models, see \textcite[chapter 6]{keelingModeling}.
% \todo{Maybe say something about still having a stochastic observation model, if not clear enough from intro}

@@

\Cref{SEIR:sec:SIR} introduces the \emph{susceptible-infected-recovered} (SIR) model, a simple compartmental model for a disease which confers immunity.
The SIR model is named because each individual in the population is divided into one of three states: susceptible, infectious, or recovered.
This model has several assumptions which make it inappropriate to model SARS-CoV-2; therefore, the remainder of this sections relaxes these assumptions.
First, \cref{SEIR:sec:non-exponential} relaxes the assumption of an exponentially-distributed infectious period to a gamma distribution.
Then, \cref{SEIR:sec:SEIR} relaxes the assumption that individuals are infectious immediately after they are infected; it does this by extending the SIR model to the SEIR model, including an \emph{exposed} compartment.
Finally, \cref{SEIR:sec:structured-populations} allows heterogeneity in the population, introducing stratification by important characteristic(s).
% Finally, \cref{SEIR:sec:time-varying-foi} relaxes the assumption of a constant transmission rate to allow for time-varying behavioural changes.

@@

\begin{figure}[h]
\makebox[\textwidth][c]{
\begin{tikzpicture}[
    node distance = 2.5cm,
    on grid,
    auto,
    ->,>=stealth',
    every state/.style={draw,rectangle},
    ]

@@

    \path (S) edge node {$\lambda(t)$} (I)
          (I) edge node {$\gamma$} (R);
\end{tikzpicture}
}
  \caption[The SIR model]{Schematic of the basic SIR model. Arrows are labelled with the rates at which individuals move between compartments.}
  \label{SEIR:fig:SIR}
\end{figure}

The simplest compartmental model for a disease which confers immunity is the (SIR) model\todo{ref for SIR being simplest}.
The name refers to the three compartments in the model, displayed in \cref{SEIR:fig:SIR}.
Individuals in the susceptible compartment are those that could be infected (they have no immunity).
The proportion of the population that is susceptible at time $t$ is $s(t)$.
Individuals in the infectious compartment can infect others, and eventually recover.
The proportion of the population that is infectious at time $t$ is $i(t)$.
Individuals in the recovered compartment are immune to the disease, and cannot be infected again.
The proportion of the population that have recovered by time $t$ is $r(t)$.
The model's state is $\vec{x}(t) = (s(t), i(t), r(t))^T$.
In the SIR model, individuals are assumed to be homogeneous, \ie they are all identical.

The SIR model is described by the following system of ODEs~\autocite[19]{keelingModeling}:
\begin{align}
\frac{ds(t)}{dt} &= -\lambda(t) s(t) \\
\frac{di(t)}{dt} &= \lambda(t) s(t) - \gamma i(t) \\
\frac{dr(t)}{dt} &= \gamma i(t)
\end{align}
where $\lambda(t) = \beta i(t)$, is the \emph{force of infection}: the risk of infection for a susceptible individual at time $t$~\autocite[17]{keelingModeling}.
Two parameters are present in this model: $\beta$ and $\gamma$, \ie $\vec{\theta} = (\beta, \gamma)^T$.
$\beta$ is the transmission rate, absorbing several terms as described below.
%\todo{check what beta is called: effective transmission rate or transmission rate or something else?}, or the number of individuals an infected individual infects per day.
$\gamma$ is the recovery rate, or the proportion of infected individuals that recover per day; hence, $1/\gamma$ is the mean infectious period~\autocite[367]{keelingModeling}.
These parameters are assumed to be constant.

% The term $\beta s(t)i(t)$ is the rate at which susceptible individuals (as a proportion of the total population size) are infected~\autocite[18]{keelingModeling}.
% Susceptible individuals move from the $s(t)$ to $i(t)$ when they are infected.
The expression for the force of infection, $\lambda(t)$ can be arrived at by considering a single infected individual~\autocite[214]{kretzschmarMathematical}.
Assume this individual, on average, comes into contact with other individuals in the population at a rate of $\kappa$ per day.
Further, assume the population is \emph{well-mixed}, meaning that each contact is with an individual chosen uniformly at random from the population; this is implied by assuming a homogeneous population.
For such a randomly selected individual, the probability that they are susceptible is $s(t)$.
% Each of the infected individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
Therefore, in expectation, the infected individual makes contact with susceptible individuals at a rate of $\kappa s(t)$.
Assume that there is a constant probability of transmission upon contact between an infected and susceptible individual, $q$.
Combined with the above, the expected rate of infections caused by this infected individual is $\kappa s(t) q$.
The two constant terms here (number of contacts and probability of transmission) are absorbed into the transmission rate, $\beta = \kappa q$.
Therefore, the rate of new infections generated by a single infectious individual is $\beta s(t)$.
Multiplying by the number of infected individuals and dividing by the population size gives the rate at which susceptible individuals are infected as a proportion of the population, $\beta s(t) i(t) = \lambda(t) s(t)$.

% The term $\gamma i(t)$, the rate at which infected individuals (as a proportion of the total population size) recover, follows directly from the definitions of $\gamma$ and $i(t)$.

A few properties and assumptions of the SIR model are worth noting.
\begin{itemize}
    \item The basic reproduction number for this model is $R_0 = \beta / \gamma$~\autocite[20]{keelingModeling}.
    Intuitively, this can be thought of as the average number of infections caused by an infectious individual per day, $\beta$, multiplied by the average number of days they are infectious for, $1/\gamma$.
    \item The effectie reproduction number for this model is $R_e(t) = \beta / \gamma s(t) = R_0 s(t)$~\autocite{pellisEstimation}.
    \item The population is closed.
    That is, the total population size is constant, and no infections are imported from outside the population.
    Births and deaths of individuals are also excluded, although they can easily be incorporated in situations where they are relevant~\autocites[26]{keelingModeling}[214]{kretzschmarMathematical}.
    \item There is no analytical solution to the system of ODEs: numerical methods are required to solve the system~\autocite[25]{keelingModeling}.
    \item In addition to the parameters, the state of the system at some time (normally the initial state, the state at time $t=0$) is required to solve the system.
    \item Immunity is assumed to last until the end of the modelled period~\autocite[61]{andersonInfectious}.
    In many contexts, including here, this assumption is justified because the modelled period is shorter than the length of immunity conferred~\autocite{milneImmunity}.
    For modelling longer time periods, waning immunity can be incorporated by allowing individuals to move from the recovered compartment back to the susceptible compartment~\autocite[40]{keelingModeling}.
\end{itemize}

\subsection{Non-exponential waiting times} \label{SEIR:sec:non-exponential}
\begin{figure}[h]
\makebox[\textwidth][c]{
\begin{tikzpicture}[
    node distance = 2.5cm,
    on grid,
    auto,
    ->,>=stealth',
    every state/.style={draw,rectangle},
    ]

@@

    \path (S) edge node {$\beta i$} (I1)
          (I1) edge node {$n\gamma$} (I2)
          (I2) edge node {$n\gamma$} (I3)
          (I3) edge node {$n\gamma$} (I4)
          (I4) edge node {$n\gamma$} (R);
\end{tikzpicture}
}
  \caption[The SIR model with non-exponential infectious period]{Schematic of the SIR model with non-exponential infectious period. The infectious period is modified by having multiple I states.}
  \label{SEIR:fig:SIR-gamma}
\end{figure}

The basic SIR model implies that the distribution of the time spent in the infected compartment is exponential~\autocite[96]{keelingModeling}.
This is an unrealistic assumption for most diseases and leads to an underestimation of the reproduction number if the model is fit to data~\autocites{lloydRealistic}{wearingAppropriate}.
A standard extension to the SIR model is to allow the infectious period to follow a gamma distribution~\autocite[94]{keelingModeling}.
This is done by adding multiple infected compartments (see \cref{SEIR:fig:SIR-gamma}).
The system can now be written as follows:
\begin{align}
\frac{ds(t)}{dt} &= -\lambda(t) s(t)\\
\frac{di_1(t)}{dt} &= \lambda(t) s(t) - n\gamma i_1(t) \\
\frac{di_2(t)}{dt} &= n\gamma i_1(t) - n \gamma i_2(t) \\
&\vdots \nonumber \\
\frac{di_n(t)}{dt} &= n\gamma i_{n-1}(t) - n \gamma i_n(t) \\
\frac{dr(t)}{dt} &= n\gamma i_n(t)
\end{align}
where $n$ is the number of infected compartments, and redefining $\lambda(t) = \beta \sum_{j=1}^n i_j(t)$.
In this model the infectious period is distributed $\GamDist(n, n\gamma)$, a realistic model for a variety of diseases~\autocite{wearingAppropriate}.
This parameterisation is chosen to give $1/\gamma$ as the mean infectious period (the gamma distribution's definition and its expectation are in \cref{E-distributions}).
% (it is slightly unclear what it means for a variable to have a distribution in a deterministic model, the clearest explanation is that this is the distribution which the ).
A gamma distribution with the first parameter equal to 1 is an exponential distribution, in which case this model collapses to the basic SIR model in \cref{SEIR:sec:SIR}.
The distribution for the time spent in the infected compartments tends to a constant (\ie the same for all individuals) as $n \to \infty$~\autocite{lloydRealistic}.
The reproduction number in terms of the parameters is unchanged by this extension, \ie $R_0 = \beta / \gamma$.

@@

\begin{figure}[h]
\makebox[\textwidth][c]{
\begin{tikzpicture}[
    node distance = 2cm,
    on grid,
    auto,
    ->,>=stealth',
    every state/.style={draw,rectangle},
    % remember picture, overlay
    ]

    \node[state, xshift=-2cm] (S) {$s$};
    \node[state, right=of S] (E1) {$e_1$};
    \node[state, right=of E1] (E2) {$e_2$};
    \node[state, right=of E2, draw=none] (E3) {$\cdots$};
    \node[state, right=of E3] (E4) {$e_m$};
    % \node[state, right=of E4] (I1) {$i_1$};
    \node[state, below=of E1] (I1) {$i_1$};
    \node[state, right=of I1] (I2) {$i_2$};
    \node[state, right=of I2, draw=none] (I3) {$\cdots$};
    \node[state, right=of I3] (I4) {$i_n$};
    \node[state, right=of I4] (R) {$r$};

    \path (S) edge node {$\beta i$} (E1)
          (E1) edge node {$m\sigma$} (E2)
          (E2) edge node {$m\sigma$} (E3)
          (E3) edge node {$m\sigma$} (E4)
          (I1) edge node {$n\gamma$} (I2)
          (I2) edge node {$n\gamma$} (I3)
          (I3) edge node {$n\gamma$} (I4)
          (I4) edge node {$n\gamma$} (R);
    \draw[->] (E4.east) -- ++(1,0) |- ([yshift=-1cm]E4.east) node [near start, right] {$m\sigma$} -|  (I1.north);
\end{tikzpicture}
}
  \caption[The SEIR model]{Schematic of the SEIR model.}
  \label{SEIR:fig:SEIR}
\end{figure}

The SIR model assumes that individuals are immediately infectious upon infection.
This is unrealistic for most diseases, as there is a period between an individual being infected and them being infectious, known as the \emph{latent period}.
\todo{I will probably end up defining the latent period in the intro, so should just ref that}
Ignoring the latent period will again overestimate the reproduction number~\autocite{wearingAppropriate}.

@@

To relax the assumption of a homogeneous population, the population can be stratified.
If the strata have differing transmission patterns, this can change the dynamics of the epidemic.
The different patterns could be due to behavioural factors (\eg high-numbers of sexual partners in a sexually-transmitted disease~\autocite[69]{keelingModeling}, younger age groups making many more contacts for respiratory diseases~\autocite[176]{andersonInfectious}), or biological factors (\eg genetics reducing the susceptibility of some individuals~\autocite[208]{andersonInfectious}).
I will assume that the disease progression (\eg length of the exposed and infectious periods) is the same for all strata, although this assumption can be relaxed.
Stratification requires two major changes model changes: extension of the state vector and describing the between-strata interactions.

The state vector is now a vector of vectors, one for each of the compartments.
I denote the set of strata, as $\set{A}$, with the set's size as $A$.
For example, in the basic SIR model, $\vec{x}(t) = (\vec{s}(t), \vec{i}(t), \vec{r}(t))^T$.
The $a$th element of each vector is the proportion of strata $a$ in that compartment.
For example, $\vec{s}(t) = (s_1(t), \dots, s_A(t))^T$ where $s_a(t)$ is the proportion of the $a$th strata that are susceptible at time $t$.
Hence, $N_a s_a(t)$ is the number of susceptible people in stratum $a$.
These vectors simply replace the scalars in the previous equations.
Various other definitions are possible and equivalent, for example defining $s_a$ as the proportion of the total population that are both in strata $a$ and susceptible or parameterised using numbers of individuals~\autocite[57]{keelingModeling}.

The interactions between strata also need describing.
This is done by replacing the scalar $\beta$ with a \emph{Who Acquires Infection From Whom} (WAIFW) matrix, $\matr{\beta}$~\autocite[58]{keelingModeling}; its interpretation is described below.
The force of infection is now a vector, $\vec\lambda(t) = \matr{\beta} \sum_{j=1}^n \vec{i_j}(t)$.
Its expression can be arrived at by considering a single infected individual in strata $a'$.
Assume any such individual comes into contact with individuals in strata $a$ at a rate of $\kappa_{aa'}$ in units of contacts per day.
% Further, assume that each contact is with an individual chosen uniformly at random from within the strata.
% For such a randomly selected individual, the probability that they are susceptible is $s_a(t)$.
% Each of the infected individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
Following the same argument as in \cref{SEIR:sec:SIR} finds that, in expectation, the infected individual makes contact with susceptible individuals in stratum $a$ at a rate of $\kappa_{aa'} s_a(t)$.
Assume that there is a constant probability of transmission upon contact between an infected in strata $a'$ and susceptible in strata $a$ individual, $q_{aa'}$.
Therefore, the expected rate of infections in strata $a$ caused by this infected individual is $\kappa_{aa'} s_a(t) q_{aa'}$.
Multiplying by the number of infected individuals in strata $a'$, $N_{a'} i_{a'}(t)$, gives $\kappa_{aa'} q_{aa'} s_a(t) N_{a'} i_{a'}(t)$.
The constant terms here are absorbed into the WAIFW matrix, giving $\beta_{aa'} = \kappa_{aa'} q_{aa'} N_{a'}$.
The total force of infection on strata $a$ from all strata is, therefore, $\lambda_a(t) = \sum_{a' \in \set{A}} \beta_{aa'} i_{a'}(t)$.
Equivalently, $\vec\lambda(t) = \matr{\beta} \sum_{j=1}^n \vec{i_j}(t)$.
If all $\beta_{aa'}$ are equal, then the model is equivalent to the homogeneous model (\ie without any structure).
The rate that individuals move between the susceptible and exposed compartments is now $\vec{s}(t) \circ \vec\lambda(t)$, where $\circ$ denotes element-wise multiplication.

In practical situations, estimating all of the $\beta_{aa'}$ parameters is infeasible, so some assumptions must be made to reduce the number of parameters~\autocite[176]{andersonInfectious}.
The infeasibility arises because the progression of the epidemic depends on $\vec{\lambda}(t)$, which has dimension $A$, while $\matr{\beta}$ has dimension $A^2$.
Common assumption include assuming symmetry, \ie $\beta_{aa'} = \beta_{a'a}$, or that the direction of transmission is irrelevant, \ie $\beta_{aa'} = \beta_{a} \beta_{a'}$.
Alternatively, as in the scalar case, $\beta_{aa'}$ can be decomposed into the contact rate between strata $a$ and $a'$, $\kappa_{aa'}$ and the probability of transmission upon contact, $q_{aa'}$, and then assumptions made about the components.
The choice of assumptions depends on the context, including the disease being studied and the data available.
